---
title: "Obesity Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Obesity Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/mrsimple07/obesity-prediction\>*

### Reference:

*\<mrsimple07. (n.d.). Obesity prediction [Data set]. Kaggle. https://www.kaggle.com/datasets/mrsimple07/obesity-prediction\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r load dataset}
# Load dataset
obesity_data <- read.csv("obesity_data.csv", colClasses = c(
  Age = "integer",
  Gender = "factor",
  Height = "numeric",
  Weight = "numeric",
  BMI = "numeric",
  PhysicalActivityLevel = "factor",
  ObesityCategory = "factor"
))

# Display the structure of the dataset
str(obesity_data)

# View the first few rows of the dataset
head(obesity_data)

# View the dataset in a separate viewer window
View(obesity_data)
```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Count of each category in Gender
gender_counts <- table(obesity_data$Gender)
print(gender_counts)

```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean, Median, and Mode of Age
mean_age <- mean(obesity_data$Age)
median_age <- median(obesity_data$Age)
mode_age <- names(sort(-table(obesity_data$Age)))[1]
print(paste("Mean Age:", mean_age))
print(paste("Median Age:", median_age))
print(paste("Mode Age:", mode_age))

```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
# Range of Height
height_range <- range(obesity_data$Height)
print(paste("Height Range:", height_range))

# Standard Deviation of Weight
weight_sd <- sd(obesity_data$Weight)
print(paste("Weight Standard Deviation:", weight_sd))

```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
# Correlation between BMI and Weight
correlation_bmi_weight <- cor(obesity_data$BMI, obesity_data$Weight)
print(paste("Correlation between BMI and Weight:", correlation_bmi_weight))

# Relationship between Gender and ObesityCategory
gender_obesity_table <- table(obesity_data$Gender, obesity_data$ObesityCategory)
print(gender_obesity_table)

```

## ANOVA
```{r ANOVA}
# Perform ANOVA
# Assuming you want to perform ANOVA to test the relationship between ObesityCategory and PhysicalActivityLevel
anova_result <- aov(BMI ~ PhysicalActivityLevel, data = obesity_data)

# Print ANOVA summary
print(summary(anova_result))

```

## Plots
```{r Plots}
# Univariate Plots
library(ggplot2)
# Histogram of Age
age_hist <- ggplot(obesity_data, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Histogram of Age", x = "Age", y = "Frequency")

# Boxplot of Weight
weight_boxplot <- ggplot(obesity_data, aes(x = "", y = Weight)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(title = "Boxplot of Weight", x = "", y = "Weight")

# Multivariate Plots
# Scatterplot of Height vs Weight colored by Gender
height_weight_scatter <- ggplot(obesity_data, aes(x = Height, y = Weight, color = Gender)) +
  geom_point() +
  labs(title = "Scatterplot of Height vs Weight by Gender", x = "Height", y = "Weight")

# Faceted histogram of BMI by ObesityCategory
bmi_hist_facet <- ggplot(obesity_data, aes(x = BMI, fill = ObesityCategory)) +
  geom_histogram(binwidth = 2, position = "dodge") +
  labs(title = "Faceted Histogram of BMI by Obesity Category", x = "BMI", y = "Frequency") +
  facet_wrap(~ObesityCategory)

# Display plots
print(age_hist)
print(weight_boxplot)
print(height_weight_scatter)
print(bmi_hist_facet)

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Check for missing values
missing_values <- sum(is.na(obesity_data))

if (missing_values > 0) {
  print("There are missing values in the dataset.")
} else {
  print("There are no missing values in the dataset.")
}

```

## Round-off Columns
```{r Round-off}
# Round off numeric columns to 2 decimal places
numeric_cols <- sapply(obesity_data, is.numeric)
obesity_data[numeric_cols] <- lapply(obesity_data[numeric_cols], round, 2)

# Print first few rows of the updated dataset
head(obesity_data)
```

# Model Training
## Data Splitting
```{r Data Splitting}
# Set seed for reproducibility
set.seed(123)

# Split data into training and testing sets
train_index <- sample(1:nrow(obesity_data), 0.7 * nrow(obesity_data))
train_data <- obesity_data[train_index, ]
test_data <- obesity_data[-train_index, ]

# Print dimensions of training and testing sets
print(paste("Training data dimensions:", dim(train_data)))
print(paste("Testing data dimensions:", dim(test_data)))
```

## Bootstrapping
```{r Bootstrapping}
# Set seed for reproducibility
set.seed(123)

# Perform bootstrapping
bootstrapped_data <- replicate(100, sample(obesity_data, replace = TRUE))

# Print dimensions of bootstrapped data
print(paste("Dimensions of bootstrapped data:", dim(bootstrapped_data)))

```

## Cross-Validation
```{r Cross-validation}
library(caret)

# Remove Age column
obesity_data <- obesity_data[, -which(names(obesity_data) == "Age")]

# Set seed for reproducibility
set.seed(123)

# Define cross-validation control
ctrl <- trainControl(method = "cv", number = 5)

# Fit model using cross-validation
cv_model <- train(ObesityCategory ~ ., data = obesity_data, method = "rpart", trControl = ctrl)

# Print cross-validation results
print(cv_model)
```

## Train Different Models
```{r Model training}
# Train CART Model
rpart_model <- train(ObesityCategory ~ ., data = obesity_data, method = "rpart", trControl = ctrl)

# Evaluate Model
print(rpart_model)

# Load necessary library
library(randomForest)

# Train Random Forest Model
rf_model <- train(ObesityCategory ~ ., data = obesity_data, method = "rf", trControl = ctrl)

# Print model
print(rf_model)

# Load necessary libraries
library(caret)
library(e1071)

# Train SVM with Radial Kernel
svm_model <- train(ObesityCategory ~ ., data = obesity_data, method = "svmRadial", trControl = ctrl)

# Print model
print(svm_model)
```

## Performance Comparison
```{r Performance Comparison}
# Compare model performance using resamples
resamples_list <- resamples(list(rpart = rpart_model, svm = svm_model, rf = rf_model))

# Summarize performance
summary_results <- summary(resamples_list)
print(summary_results)
```

# Saving Model
```{r Saving Model}
# Saving the Random Forest model
saveRDS(rf_model, "./models/saved_rf_model.rds")

# Load the saved model
loaded_rf_model <- readRDS("./models/saved_rf_model.rds")

# Prepare new data for prediction
new_data <- data.frame(
  Gender = "Male",
  Height = 175,
  Weight = 80,
  BMI = 26,
  PhysicalActivityLevel = "3"
)

# Use the loaded model to make predictions
predictions_loaded_model <- predict(loaded_rf_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```

